---
title: "P8015_hw3_dw3093"
author: "Katherine Wang"
date: "2024-10-13"
output: github_document
---

```{r setup, include=FALSE}
  library(p8105.datasets)
  library(dplyr)
  library(ggplot2)
  library(ggridges)
  library(tidyverse)
  library(patchwork)
knitr::opts_chunk$set(
	echo = TRUE,
	warning = FALSE,
	fig.width = 8, 
  fig.height = 6,
  out.width = "90%"
)

theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```

## Question 1 ##
```{r include=FALSE}
data("ny_noaa")
ny_noaa |> 
  summary()
na_counts = colSums(is.na(ny_noaa))
print(na_counts)
```
Summary of the dataset
This dataset contains `nrow(ny_noaa)` rows and `ncol(ny_noaa)` columns. 
This dataset contains 2,595,176 rows and 7 columns
The date of observation, ranging from 1981-01-01 to 2010-12-31.

Key variables include:
- weather station ID (`id`)
- date of observation (`date`)
- precipitation in tenths (mm) (`prcp`)
- snowfall (mm) (`snow`)
- snow depth (mm)  (`snwd`)
- maximum temperature (tenths of degrees C) (`tmax`)
- minimum temperature (tenths of degrees C) (`tmin`)

The `tmax` and `tmin` are set to be character variables, so they need to converse to numeric variables in the future step.

Through the missing value check, there are a great amount missing (NAs) values as shown on the table:
     id    date    prcp    snow    snwd    tmax    tmin 
      0       0  145838  381221  591786 1134358 1134420 

```{r echo=FALSE}
ny_noaa = 
  ny_noaa|>
  mutate(
    tmax = as.numeric(tmax),
    tmin = as.numeric(tmin)
    )
summary(ny_noaa)
```
Convert tmax and tmin to numeric variables.

```{r echo=FALSE}
ny_noaa = ny_noaa |>
  mutate(
    year = year(date),
    month = month(date),
    day = day(date),
    tmax = tmax / 10,  
    tmin = tmin / 10,   
    prcp = prcp / 10,   
  )

com_snowfall = ny_noaa |> 
  count(snow) |> 
  arrange(desc(n))

print(com_snowfall)
```
The most commonly observed value for snowfall is 0, which indicates that no snowfall occurred on the majority of days in this dataset that is 2,008,508 observations. This is followed by a large number of NAs, totaling 381,221 observations, which suggests that snowfall data was not recorded or available for these entries. Most days in New York State do not experience snowfall, especially during non-winter months. Therefore, it is expected that the majority of observations would record 0 for snowfall, representing days without snow. 

```{r}
ny_noaa |>
  group_by(id,year,month)|>
  filter(month %in% c(1, 7)) |>
  mutate(
    month = factor(month(date), 
                   levels = c(1, 7), 
                   labels = c("January", "July"))
    ) |>
  summarise(avg_tmax = mean(tmax, na.rm = TRUE))|>
  ggplot(aes(x = year, y = avg_tmax))+
  geom_point(alpha = .1)+
  geom_line()+
  facet_grid(.~month)+
  labs( title = "Avg. Max Temperature in January and July across Years by Station",
    x = "Year",
    y = "Avg. Max Temperature (°C)")+ 
  theme_minimal() + 
  theme(legend.position = "bottom")
```
The two-panel plot shows that avg. max temperatures in January exhibit more variability across stations and years ranging from about -10°C to 10°C, while July temperatures are relatively stable, consistently around 25°C to 35°C. 
In January more fluctuations are visible, with few potential outliers like the lowest temperature is about 13°C in around 1982. July shows minimal variation, the temperature is quite consistency warm across the years, but still contains few outliers, like the lowest temperature is about 14°C in around 1988.


```{r}
hexbin_plot=
  ny_noaa|>
  ggplot(aes(x = tmin, y = tmax))+
  geom_hex()+
  labs(
    x = "minimum temperature(°C)",
    y = "maximum temperature(°C)",
    )+
    theme_minimal()

density_plot = 
  ny_noaa|>
  filter(snow>0, snow<100)|>
  ggplot(aes(x = snow, y = as.factor(year))) + 
  geom_density_ridges()

combine_plot =
  hexbin_plot + density_plot

print(combine_plot)
```
The hexbin plot shows a strong positive relationship between tmax and tmin, with most data points clustered along a diagonal, indicating that higher minimum temperatures are associated with higher maximum temperatures. 
The ridge plot reveals that most snowfall values fall below 50 mm, with distinct peaks around 25 mm and 75 mm, and some variation in snowfall distributions across years, though the overall pattern remains consistent.

## Question 2 ##
#load, clean, and tidy 'nhanes_covar' dataset
```{r include=FALSE}
covar = 
  read_csv("~/P8105_hw3_dw3093/HW3Q2data/nhanes_covar.csv", 
           skip = 4, 
           na = c("NA",".", "")
           )|>
  janitor::clean_names()|>
  filter(age >= 21)|>
  drop_na()|>
  mutate(
    sex = factor(
      sex,
      levels = c(1,2),
      labels = c("Male", "Female")),
  education = factor(
    education,
    levels = c(1,2,3),
    labels = c("Less than high school",
               "High school equivalent",
               "More than high school")))
```

#load, clean, and tidy 'nhanes_accel' dataset
```{r include=FALSE}
accel = 
  read_csv("~/P8105_hw3_dw3093/HW3Q2data/nhanes_accel.csv", 
           na = c("NA",".", "")
           )|>
  janitor::clean_names()
```

#merge 'nhanes_covar' and 'nhanes_accel' dataset to 'merged_df'
```{r}
merged_df =
  left_join(covar, accel, by = "seqn")
```
After the data cleaning and merging, the merged_df have 228 observation and 1445 variables.

#create a reader-friendly table for the number of men and women in each education category
```{r}
merged_df |>  
  group_by(sex, education) |>
  summarise(count = n(), .groups = "drop") |>
  pivot_wider(
    names_from = sex,
    values_from = count
    ) |>
  knitr::kable()
```
The reader-friendly table shows that 27 males and 28 females have less than a high school education, 35 males and 23 females have a high school equivalent education, and 56 males and 59 females have more than a high school education. 

For the "less than a high school education" and "more than a high school education" categories, the number between males and females is relatively balanced, but for the "high school equivalent education" category, there are much fewer males than females.

#create a visualization of the age distributions for men and women in each education category
```{r}
merged_df |>
  ggplot(aes(x = sex, y = age, fill = education)) +
  geom_boxplot()+
  facet_grid(. ~ education) +
  labs(title = "Age Distribution by Gender and Education", 
       x = "Sex", 
       y = "Age (year)")+
   theme_minimal()
```
The box plot shows that participants with less than a high school education are older, with both males and females having a similar median age around 60 years and a narrow age range. In the high school equivalent category, males tend to have a slightly higher median age compared to females, and both genders show a broader age range, with males having a particularly widespread from their 20s to 80s. In the more than high school category, males and females have similar age distributions, both with a median age of around 60 years, though females exhibit a slightly broader range. 

# Plot total activities vs. age
```{r}
merged_df |> 
  mutate(
    total_activity = rowSums(across(starts_with("min")), na.rm = TRUE)
  ) |>
  select(seqn, sex, age, bmi, education, total_activity) |> 
  ggplot(aes(x = age, y = total_activity, color = sex)) + 
  geom_point(alpha = .5) + 
  geom_smooth(se = FALSE) + 
  facet_wrap(~education) + 
  labs(
    title = "Total Activity vs. Age by Gender and Education Level", 
    x = "Age", 
    y = "Total Activity"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom"
  )
```

- In the "less than high school" group, females initially have higher activity levels, but their activity sharply declines after age 60, eventually falling below that of males, who show a more gradual decline with age. 
- In the "high school equivalent" group, females are more active than males at younger ages, but their activity decreases more steeply after age 50, converging with male levels by age 60. Males exhibit a steadier decline across age in this group. 
- In the "high school equivalent" and "more than high school" groups, females have notably higher activity in their 20s and 30s but diminish by age 60, where male and female activity levels converge and decline together. 
- Overall, females tend to be more active in their younger years, while males experience more gradual decreases in activity, often maintaining higher levels in older age, particularly in the lower education groups, which indicates that both gender and education influence activity trends as people age.

#create a three-panel plot that shows the 24-hour activity time courses for each education level
```{r}
hourly_act <- merged_df |>
  pivot_longer(
    cols = starts_with("min"),
    names_prefix = "min",
    names_to = "minute",
    values_to = "activity_level"
  ) |>
  mutate(minute = as.numeric(minute),
         hour = minute / 60)

ggplot(hourly_act, aes(x = hour, y = activity_level, color = sex)) +
  geom_line(stat = "summary", fun = "mean") +
  geom_smooth(se = FALSE) +
  facet_wrap(~education) +
  scale_x_continuous(breaks = seq(0, 24, by = 3), 
                     labels = c("12", "3", "6", "9", "12", "15", "18", "21", "24")) +
  labs(
    title = "24-Hour Activity Time Courses by Education and Gender", 
    x = "Hour of the Day (24-hour Clock)", 
    y = "Avg. Activity Level"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```
The graph shows 24-hour activity patterns for men and women across different education levels. Across both male and female groups, there is a sharp increase in activity in the morning (around 6-9), peaking between 9 and 12 am, followed by a gradual decline in the afternoon and a smaller peak in the afternoon (around 15-18). Activity levels significantly drop off after 21 and remain low during nighttime hours. Females consistently show higher average activity levels compared to males throughout the day, especially during the peak activity periods. While the overall patterns are similar across education levels, individuals with less than a high school education tend to have slightly lower activity levels, and those with higher education exhibit more sustained and pronounced morning peaks. 
In conclusion: the trends suggest that both gender and education influence daily activity patterns, with females generally being more active and higher education associated with slightly higher and longer-lasting activity peaks during the day.

## Question 3 ##
Import and tidy 4 datasets
```{r}
citi_jan20 =
  read.csv("~/P8105_hw3_dw3093/HW3Q3data/Jan 2020 Citi.csv", na = c("NA", ".","")) |>
  janitor::clean_names() |>
  mutate(
    month = "January",
    year = 2020
  )

citi_jan24 =
  read.csv("~/P8105_hw3_dw3093/HW3Q3data/Jan 2024 Citi.csv", na = c("NA", ".","")) |>
  janitor::clean_names() |>
  mutate(
    month = "January",
    year = 2024
  )

citi_jul20 =
  read.csv("~/P8105_hw3_dw3093/HW3Q3data/July 2020 Citi.csv", na = c("NA", ".","")) |>
  janitor::clean_names() |>
  mutate(
    month = "July",
    year = 2020
  )

citi_jul24 =
  read.csv("~/P8105_hw3_dw3093/HW3Q3data/July 2024 Citi.csv", na = c("NA", ".","")) |>
  janitor::clean_names() |>
  mutate(
    month = "July",
    year = 2024
  )

citi_merged = 
  bind_rows(citi_jan20, citi_jan24,citi_jul20, citi_jul24)|> 
  relocate(month, year)
```
I imported 4 datasets from Citi Bike data files for January 2020, July 2020, January 2024, and July 2024. Then I combined all four datasets into one and organized the columns by moving month and year to the front.
The merged dataset contains 99,485 observations and 9 key variables:
1. month, 2. year, 3. ride_id, 4. rideable_type, 5. weekdays, 6. duration, 7. start_station_name, 8. end_station_name, 9. member_casual

#Produce a reader-friendly table showing the total number of rides in each combination of year and month separating casual riders and Citi Bike members
```{r}
citi_merged|> 
  group_by(year, month, member_casual) |> 
  summarize(
    total_rides = n()) |>
  pivot_wider(
    names_from = member_casual,
    values_from = total_rides
  )|>
  arrange(year, month) |>
  knitr::kable()
```
The created reader-friendly table shows that Citi Bike membership usage consistently outpaces casual ridership, and both groups exhibit higher ride activity in the summer months, specifically:
In January 2020, casual riders took 984 rides, while members took 11,436 rides, showing significantly higher usage by members.
In July 2020, the number of rides increased for both groups, with casual riders taking 5,637 rides and members taking 15,411 rides.
In January 2024, the number of rides increased further, with casual riders taking 2,108 rides and members taking 16,753 rides.
By July 2024, there had been a sharp rise in rides, with casual riders accounting for 10,894 rides and members for 36,262 rides. This reflected a dramatic increase in popularity (maybe due to post-COVID people's need to commute or gather).

#Make a table showing the 5 most popular starting stations for July 2024; include the number of rides originating from these stations.
```{r}
citi_jul24 |> 
   group_by(start_station_name) |> 
  summarise(
    total_rides = n()
    ) |> 
  mutate(rank = dense_rank(desc(total_rides))) |>  
  filter(rank <= 5) |>
  arrange(desc(total_rides)) |> 
  knitr::kable()
```
The 5 most popular starting stations for July 2024 table shows that Pier 61 at Chelsea Piers saw the highest number of rides, 163. It is followed by University Pl & E 14 St, with 155 rides, W 21 St & 6 Ave, with 152 rides, and West St & Chambers St, with 150 rides. The fifth most popular station is W 31 St & 7 Ave, with 146 rides.

#Make a plot to investigate the effects of day of the week, month, and year on median ride duration
```{r}

```


